---
title: "Pivoting Table with R"
author: "JM Muriuki"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
  latexengine: default
  html_document: default
---

## attach libraries
```{r}
library(tidyverse)
library(readxl)
library(here)
library(skimr) # install.packages('skimr')
library(kableExtra) # install.packages('kableExtra')

```

## getting to know my directory
```{r}
getwd()
```


## read in data
```{r}
lobsters <- read_excel(here("lobsters.xlsx"),skip=4)
```
## pivoting a data 
```{r}
#data %>% group_by() %>% summarize()
```

## group_by one variable
```{r}
lobsters %>% group_by(year) %>% summarize(count_by_year = n())
```
### when you don't group_by first?
```{r}
lobsters %>%  summarise(count=n())
```
### what if we only group_by?
```{r}
lobsters %>% group_by(year)
```
## group_by multiple variables
```{r}
lobsters %>%  group_by(site,year) %>% summarise(count_by_siteyear=n())
```
## summarize multiple variables
```{r}
lobsters %>% group_by(site, year) %>% summarize(count_by_siteyear = n() ,mean_size_mm = mean(size_mm))
# some of the means are  passed as NA because one or more   values in that year are NA.

lobsters %>% group_by(site, year) %>% summarize(count_by_siteyear =  n(), mean_size_mm = mean(size_mm, na.rm=TRUE), sd_size_mm = sd(size_mm, na.rm=TRUE))
```
## Adding a variable assignment to that first line:
```{r}
siteyear_summary <- lobsters %>% group_by(site, year) %>% summarize(count_by_siteyear =  n(), mean_size_mm = mean(size_mm, na.rm = TRUE), sd_size_mm = sd(size_mm, na.rm = TRUE))
```
### inspect our new variable
```{r}
siteyear_summary
```
## Table formatting with kable()
### kable() offers a nice presentation of our table
```{r}
## make a table with our new variable
siteyear_summary %>% kable()
```
## Activity
### Building from our analysis
```{r}
siteyear_summary <- lobsters %>% group_by(site,year) %>% summarise(count_by_siteyear= n(),mean_size_mm=mean(size_mm,na.rm = TRUE),median_size_mm=median(size_mm,na.rm=TRUE))

```
### a ggplot option:
```{r}
ggplot(data = siteyear_summary,aes(x=year,y=median_size_mm,color=site))+geom_line()
```
### save the image
```{r}
ggsave(here("figures","lobsters-line.png"))
```
### another option
```{r}
ggplot(siteyear_summary,aes(x=year,y=median_size_mm))+geom_col()+facet_wrap(~site)
```
### save here
```{r}
ggsave(here("figures","lobsters-col.png"))
```
## Oh no! They sent me the wrong data
```{r}
# now the data becomes
lobsters <- read_xlsx(here("lobsters2.xlsx"),skip=4)
# Then the same procedure follows
```
## mutate()

###There are a lot of times where you don't want to summarize your data, but you do want to operate beyond the original daa.This is often done b y adding a column.
```{r}
lobsters %>% mutate(size_m=size_mm/1000)
```
### Adding a column that has the same value repeated
```{r}
lobsters_detailed <- lobsters %>% mutate(size_m=size_mm/1000,millenia=2000,observer="Allison Horst")
```
## select()
```{r}
lobsters_detailed %>% select(date,site,size_m)
```






